---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r}
act_data <- read.csv("activity.csv", header = TRUE, sep = ",")
act_data$date <- strptime(act_data$date, "%Y-%m-%d") # transform to a date
str(act_data)
summary(act_data)
head(act_data)

```

## What is mean total number of steps taken per day?

Remove the NAs from the data set and aggregate steps per day

```{r}
## load package
library(dplyr)
## remove NAs from data set & aggregate steps per day
num_day_data <- act_data %>%
      group_by(date) %>%
      filter(steps != is.na(steps)) %>%
      summarise(daily_steps = sum(steps))
## get insight into the resulting data set
str(num_day_data)
summary(num_day_data)
head(num_day_data)
```

Histogram of total of steps per day without NAs

```{r}
plot(num_day_data, type = "h", ylab = "number of steps", xlab = "date", main = "total steps by day")
```

Calcuting the mean and median
```{r}
mean_daily <- as.integer(mean(num_day_data$daily_steps))
median_daily <- as.integer(median(num_day_data$daily_steps))
```

Mean = `r mean_daily`

Median = `r median_daily`

## What is the average daily activity pattern?

Create a time series line plot with x-axis 5 minute interval and y-axis 
```{r}
## remove NAs from data set & averge steps per interval
num_interval_data <- act_data %>%
      group_by(interval) %>%
      filter(steps != is.na(steps)) %>%
      summarise(interval_steps = mean(steps))
## get insight into the resulting data set
str(num_interval_data)
summary(num_interval_data)
head(num_interval_data)

plot(num_interval_data$interval, num_interval_data$interval_steps, type = "l")
```

Determine the 5 minute interval with the highest average steps

```{r}
max_value <- max(num_interval_data$interval_steps)
max_interval <- num_interval_data$interval[num_interval_data$interval_steps == max_value]
```

Highest average steps in any interval = `r max_value`

The interval with the highest average steps = `r max_interval`

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum_missing_values <- sum(is.na(act_data$steps))  # for the steps data
```

Number of rows that have missing values for steps: `r sum_missing_values`

### Create a new dataset that is equal to the original dataset but with the missing data filled in. 

Strategy: Filling in the missing values with mean of that interval across the period

```{r}
## remove NAs from data set by replacing the NA's
imputed_data <- act_data %>%
      mutate(new_steps = ifelse(is.na(steps), num_interval_data$interval_steps[num_interval_data$interval == interval], steps) )

## get insight into the resulting data set
str(imputed_data)
summary(imputed_data)
head(imputed_data)
```

### Histogram
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
num_day_imputed_data <- imputed_data %>%
      group_by(date) %>%
      summarise(daily_steps = sum(new_steps))
## get insight into the resulting data set
str(num_day_imputed_data)
summary(num_day_imputed_data)
head(num_day_imputed_data)

plot(num_day_imputed_data, type = "h", ylab = "number of steps (with imputed data)", xlab = "date", main = "total steps by day")

```

Calcuting the mean and median

```{r}
# mean and median calculated with imputed data
mean_imputed_daily <- as.integer(mean(num_day_imputed_data$daily_steps)) 
median_imputed_daily <- as.integer(median(num_day_imputed_data$daily_steps))

# difference with original data
div_mean <- (1 - mean_imputed_daily / mean_daily)
div_median <- (1 - median_imputed_daily / median_daily)
```

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean = `r mean_daily`

Mean = `r mean_imputed_daily`

Change Mean = `r div_mean`

Median = `r median_daily`

Median = `r median_imputed_daily`

Change Median = `r div_median`

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# using data set "imputed_data" as source

category_var <- as.factor(c("weekday", "weekend"))
print(category_var)
str(category_var)

weekenddays <-c("Sunday", "Saturday")

imputed_data_categorised <- imputed_data %>%
      mutate(day_of_week = weekdays(date)) %>%
      #mutate(category = ifelse((day_of_week %in% weekenddays), as.factor(category_var[2]), as.factor(category_var[1])))
      mutate(category = factor(ifelse((day_of_week %in% weekenddays), "weekend", "weekday")))

str(imputed_data_categorised)
summary(imputed_data_categorised)
head(imputed_data_categorised)
```


